Anomaly detection with kernel density estimation on manifolds

### 2022-07-05 paper/paper.Rmd
Move swiss roll to appendix b

# Kernel density estimation on manifold learning embeddings {#vkde}

@@

# Distotion-corrected kernel density estimator (DC-KDE) {#dckde}

However, in application, we usually found the observations lie on a $d$-dimensional manifold embedded in a $s$-dimensional space, where $d \ll s$. Then manifold learning can be applies to find the $d$-dimensional representation of the manifold $M$. Therefore, based on the estimator in \@ref(eq:denriem), we propose a kernel density estimator of $f$ with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be 
\begin{equation}
\label{eq:denestimator}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
\end{equation} 
where $h_i, i = 1, 2, \dots, N$ is the pointwise variable bandwidth matrix and $h_i$ is positive definite with dimension $d\times d$.

@@

### Riemannian metric as variable bandwidth

@@

# Outlier detection on manifold learning embedding

@@

### Swiss roll mapping

@@

Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space. One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \@ref(eq:swissroll). The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in the left plot of \autoref{fig:mappings}. The four colors in the mappings represent the four Gaussian kernels used to generate the meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$. 
\begin{equation}
\label{eq:swissroll}
\left\{
\begin{array}{lcl}
X = X_1 \cos{X_1}, \\
Y = X_2, \\
Z = X_1 \sin{X_1}.
\end{array}
\right.
\end{equation}

### Twin peaks mapping

For comparison, we use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks
mapping in Equation \@ref(eq:twinpeak), with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}.
\begin{equation}
\label{eq:twinpeak}
\left\{ 
\begin{array}{lcl}
X = X_1, \\
Y = X_2, \\
Z = \sin(\pi X_1) \tanh (3 X_2).
\end{array}
\right.
\end{equation}

@@

\clearpage\appendix

@@

@@

<!-- ### Isometry -->

@@

## Volume density function [MOVE TO SECTION 2.2.1]

For any tangent vector $u \in T_pM$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$, for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$. Then $Dw_i/dt(0)$ forms an orthonomal basis of the tangent space $T_pM$. Define the volume density function on $M$, $\theta_p(q): T_pM \rightarrow \mathbb{R}$, as 
$$
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
$$ 
<!-- $$ -->
<!-- \theta_p: q \rightarrow \theta_p(q)=\frac{\mu_{\exp_p{g}}}{\mu_{g_p}}(\exp^{-1}_p(q)), -->
<!-- $$ --> 
<!-- Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.  -->
The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, and $\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates 
$$
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
$$
The volume density function is motivated by how a measure $du$ changes under the change of coordinates. It is related to the integration on manifolds and plays an important part in studing the density of manifolds.

@@

## Example of $S^1$ for manifold density estimation

@@

<!-- Polar coordinates: -->
<!-- $\phi(\exp_p{vt})^{\prime}(0) = (\alpha + vt)^\prime\bigg|_{t=0} = v$. -->

<!-- Cartesian coordinates: -->
<!-- $$ -->
<!-- \psi(\exp_p{vt})^{\prime}(0) = (\cos(\arccos(p)+vt))^\prime\bigg|_{t=0} \\ -->
<!-- = (p \cos{v} - \sqrt{1-p^2} \sin{v})^\prime \bigg|_{t=0} \\ -->
<!-- = - \sqrt{1 - p^2}v -->
<!-- $$ -->
<!-- Then the volume density function is -->
<!-- $$ -->
<!-- \theta_p(q) = \sqrt{(-\sqrt{1-p^2})^2} = \sqrt{1-p^2}, -->
<!-- $$ -->
<!-- and the geodesic distance is -->
<!-- $$ -->
<!-- d_g(p,q) = \arccos(p) - \arccos(q). -->
<!-- $$ -->

>>>>

# Kernel density estimate on Riemannian manifold embeddings {#vkde}

@@

## Distotion-corrected kernel density estimator (DC-KDE) {#dckde}

However, in application, we usually find the observations lie on a $d$-dimensional manifold embedded in a $s$-dimensional space, where $d \ll s$. Then manifold learning can be applies to find the $d$-dimensional representation of the manifold $M$. Therefore, based on the estimator in \@ref(eq:denriem), we propose a kernel density estimator of $f$ with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be 
\begin{equation}
\label{eq:denestimator}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
\end{equation} 
where $h_i, i = 1, 2, \dots, N$ is the pointwise variable bandwidth matrix and $h_i$ is positive definite with dimension $d\times d$.

@@

## Riemannian metric estimation

@@

## Outlier detection on manifold learning embedding

@@

@@

Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space. We use the same 2-D meta data in \autoref{fig:metadensity} with a different mapping function, twin peaks
mapping in Equation \@ref(eq:twinpeak), with the corresponding 3-D structure shown in the right plot of \autoref{fig:mappings}. For simplicity, the simulation results for the swiss roll mapping in te left plot is shown in the Appendix \autoref{swissrollappe}.
\begin{equation}
\label{eq:twinpeak}
\left\{ 
\begin{array}{lcl}
X = X_1, \\
Y = X_2, \\
Z = \sin(\pi X_1) \tanh (3 X_2).
\end{array}
\right.
\end{equation}

@@

<!-- \clearpage -->

@@

\appendix

@@

<!-- ## Isometry -->

@@

<!-- ## Volume density function [MOVED TO SECTION 2.2]

For any tangent vector $u \in T_pM$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$, for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$. Then $Dw_i/dt(0)$ forms an orthonomal basis of the tangent space $T_pM$. Define the volume density function on $M$, $\theta_p(q): T_pM \rightarrow \mathbb{R}$, as 
$$
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
$$

The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, and $\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates 
$$
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
$$
The volume density function is motivated by how a measure $du$ changes under the change of coordinates. It is related to the integration on manifolds and plays an important part in studing the density of manifolds. -->

@@

## Example of $S^1$ for manifold density estimation [TODO: rewrite as appendix B with plots?]

@@

\clearpage

# Appendix: Simulation with swiss roll mapping {#swissrollappe}

In this appendix, we demonstrate the simulation results for the data in \autoref{twodgaussian} with the swiss roll mapping.

One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \@ref(eq:swissroll). The two-dimensional meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$ is transformed into the three-dimensional data $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$, shown in the left plot of \autoref{fig:mappings}. The four colors in the mappings represent the four Gaussian kernels used to generate the meta data $(\pmb{X}_1, \pmb{X}_2)^\prime$. 
\begin{equation}
\label{eq:swissroll}
\left\{
\begin{array}{lcl}
X = X_1 \cos{X_1}, \\
Y = X_2, \\
Z = X_1 \sin{X_1}.
\end{array}
\right.
\end{equation}
